---
title: "Demographic Data"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook contains the code for creating neighborhood population estimates.

## Dependencies
This notebook requires:

```{r load-packages}
# tidyverse packages
library(dplyr)         # data wrangling
library(readr)         # working with csv data

# spatial packages
library(areal)         # interpolation
library(sf)            # working with spatial data
library(tidycensus)    # census api access
library(tigris)        # tiger/line api access

# other packages
library(here)          # file path management
```

## Download Data
Census data will be used and is accessed via the Census Bureau's API. The `tidycensus` and `tigris` packages are used for this.

First, we'll download a shapefile of the city's census geography:

```{r get-tracts}
tracts(state = 29, count = 510, class = "sf") %>%
  select(GEOID) -> tracts
```

Next, we'll download the 2017 five year population estimates, and merge them with the census geography data we've already cleaned above:

```{r get-pops}
# 2017
get_acs(year = 2017, geography = "tract", variable = "B01003_001", 
        state = 29, county = 510) %>%
  left_join(tracts, ., by = "GEOID") %>%
  st_transform(crs = 102696) -> tracts17
```

## Load Neighborhood Data
To estimate populations for neighborhoods, we'll use the city's neighborhood boundary data:

```{r load-neighborhoods}
st_read(here("data", "raw", "nhood", "STL_BOUNDARY_Nhoods.shp"), stringsAsFactors = FALSE) %>%
  st_transform(crs = 102696) %>%
  select(NHD_NUM, NHD_NAME) -> nhoods
```

## Estimate Neighborhood Populations
We'll use a technique called areal weighted interpolation to estimate populations for each neighborhood.

```{r interpolate}
aw_interpolate(nhoods, tid = NHD_NUM, source = tracts17, sid = GEOID, 
               weight = "sum", output = "tibble", extensive = "estimate") %>%
  left_join(nhoods, ., by = "NHD_NUM") %>%
  rename(pop17 = estimate) -> nhoods
```

## Clean-up Enviornment
We can get rid of the `tracts` and `tracts17` objects at this stage.

```{r rm-tracts}
rm(tracts, tracts17)
```

## Add Part 1 Crime Counts by Neighborhood
Next, we want to add counts of crimes for each neighborhood. To do this, we need to load the crime data we cleaned previously:

```{r load-crimes}
p1_crimes <- read_csv(here("data", "clean", "p1_crimes.csv"))
```

Then, we need to project the data, place them in the appropriate coordinate system, and perform a spatial join to identify the neighborhood each crime occurs in:

```{r crimes-by-neighborhood}
# obtain counts by neighborhood
p1_crimes %>%
  st_as_sf(coords = c("XCoord", "YCoord"), crs = 102696) %>%
  st_intersection(., nhoods) %>%
  group_by(NHD_NUM) %>%
  summarise(p1Crimes = n()) -> nhoodCrimes

# remove geometry
st_geometry(nhoodCrimes) <- NULL

# join with original neighborhood data
nhoods <- left_join(nhoods, nhoodCrimes, by = "NHD_NUM")
```

Since crime rates are the most important metric, we'll add a measure of crimes per 1,000 residents in each neighborhood:

```{r add-crime-rates}
nhoods <- mutate(nhoods, p1Rate = (p1Crimes/pop17)*1000)
```

## Add Violent Crime Counts by Neighborhood
Next, we want to add counts of violent crimes for each neighborhood. To do this, we need to load the crime data we cleaned previously:

```{r load-violent-crimes}
v_crimes <- read_csv(here("data", "clean", "v_crimes.csv"))
```

Then, we need to project the data, place them in the appropriate coordinate system, and perform a spatial join to identify the neighborhood each crime occurs in:

```{r violent-crimes-by-neighborhood}
# obtain counts by neighborhood
v_crimes %>%
  st_as_sf(coords = c("XCoord", "YCoord"), crs = 102696) %>%
  st_intersection(., nhoods) %>%
  group_by(NHD_NUM) %>%
  summarise(vCrimes = n()) -> nhoodVCrimes

# remove geometry
st_geometry(nhoodVCrimes) <- NULL

# join with original neighborhood data
left_join(nhoods, nhoodVCrimes, by = "NHD_NUM") %>%
  mutate(vCrimes = ifelse(is.na(vCrimes) == TRUE, 0, vCrimes)) -> nhoods
```

Since crime rates are the most important metric, we'll add a measure of crimes per 1,000 residents in each neighborhood:

```{r add-violent-crime-rates}
nhoods <- mutate(nhoods, vRate = (vCrimes/pop17)*1000)
```

## Write Data
Finally, we'll write our neighborhood data to a shapefile for mapping:

```{r write-data}
st_write(nhoods, 
         here("data", "clean", "STL_CRIME_Neighborhoods", "STL_CRIME_Neighborhoods.shp"),
         delete_dsn = TRUE)
```
