---
title: "Density Map - Part 1 Crimes"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook contains the code for creating density maps of crimes in St. Louis:

## Dependencies
This notebook requires:

```{r load-packages}
# tidyverse packages
library(dplyr)         # data wrangling
library(readr)         # working with csv data

# spatial packages
library(sf)            # working with spatial data
library(tigris)        # tiger/line api access        
library(tmap)          # mapping
library(tmaptools)     # mapping

# other packages
library(here)          # file path management
```

## Load Data
We'll load both the crime data and the neighborhood boundaries for mapping:

```{r load-data}
read_csv(here("data", "clean", "p1_crimes.csv")) %>%
  st_as_sf(coords = c("XCoord", "YCoord"), crs = 102696) -> crimes

st_read(here("data", "clean", "STL_CRIME_Neighborhoods", "STL_CRIME_Neighborhoods.shp")) %>%
  st_transform(crs = 102696) -> nhoods
```

## Download Data
Next, we need to get St. Louis's boundary for both mapping and some final data cleaning:

```{r download-county}
counties(state = 29, class = "sf") %>%
  select(COUNTYFP) %>%
  filter(COUNTYFP == 510) %>%
  st_transform(crs = 102696) -> stLouis
```

## Removing Crimes Outside of St. Louis
We don't want to map crimes that happen in the county or elsewhere, so we'll complete a spatial join and remove crimes that do not fall within the city:

```{r clip}
st_intersection(crimes, stLouis) %>%
  select(-xyStatus, -COUNTYFP) -> crimes_stl
```

## Density Map
To create a density map, we need to specify breaks for our values:

```{r calculate-breaks}
crime_densities <- smooth_map(crimes_stl, bandwidth = 0.5, 
  breaks = c(0, 25, 50, 100, 200, 300), cover = stLouis)
```

We can then create our map:

```{r create-map}
tm_shape(crime_densities$polygons) +
  tm_fill(col = "level", palette = "BuPu", 
    title = expression("Crimes per " * km^2)) + 
  tm_shape(stLouis) + 
  tm_borders() +
  tm_shape(nhoods) + 
  tm_borders(col = "grey25", alpha = .5) +
  tm_style("gray") +
  tm_credits("Map by Christopher Prener, Ph.D. \nData via SLMPD for 2016-2017", position = c("left", "BOTTOM")) +
  tm_layout(
    main.title = "Part 1 Crimes in St. Louis",
		main.title.position = "left",
		attr.outside = TRUE, 
		legend.outside = TRUE,
		inner.margins = c(.05, .05, .05, .05)
		) -> densityMap

densityMap
```

This looks roughly like the map produced in the Belmar proposal.

## Write Map
We'll save the map for publication:

```{r write-map}
tmap_save(densityMap, filename = here("results", "density_map_p1.png"), 
          dpi = 500)
```
