--
title: "Population Plot"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook contains the code for creating a plot showing the relationship between neighborhood population and number of crimes.

## Dependencies
This notebook requires:

```{r load-packages}
# tidyverse packages
library(ggplot2)       # data plotting

# spatial packages
library(sf)            # working with spatial data

# other packages
library(here)          # file path management
```

## Load Data
We'll load the neighborhood boundaries for plotting:

```{r}
st_read(here("data", "clean", "STL_CRIME_Neighborhoods", "STL_CRIME_Neighborhoods.shp")) %>%
  st_transform(crs = 102696) -> nhoods
```

## Create Part 1 Crime Plot
First, we'll create our scatterplot object for part 1 crimes.

```{r part1-plot}
ggplot(data = nhoods, mapping = aes(x = pop17, y = p1Crimes)) +
  geom_point(position = "jitter") +
  geom_smooth(method = "lm") +
  labs(
    title = "Population and Part 1 Crime in St. Louis",
    subtitle = "By Neighborhood, 2016-2017",
    x = "Population Count, 2017",
    y = "Count of Part 1 Crimes",
    caption = "Plot by Christopher Prener, Ph.D."
  )
```

We'll write the plot to the `results/` folder for publication:

```{r write-plot}
ggsave(filename = here("results", "pop_plot_p1.png"), dpi = 500)
```

## Create Violent Crime Plot
Next, we'll create our scatterplot object for violent.

```{r violent-plot}
ggplot(data = nhoods, mapping = aes(x = pop17, y = vCrimes)) +
  geom_point(position = "jitter") +
  geom_smooth(method = "lm") +
  labs(
    title = "Population and Violent Crime in St. Louis",
    subtitle = "By Neighborhood, 2016-2017",
    x = "Population Count, 2017",
    y = "Count of Violent Crimes",
    caption = "Plot by Christopher Prener, Ph.D."
  )
```

We'll also write the plot to the `results/` folder for publication:

```{r write-violent-plot}
ggsave(filename = here("results", "pop_plot_v.png"), dpi = 500)
```
